---
title: "Nanopore pipeline"
author: "Ayesha"
date: "2025-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Install prerequisites for the quality control, alignment and preprocessing. 
Note: install and execute these in base conda environment and then use analysis tools in QIIME2 to avoid dependency conflicts 

```{r cars}
 # Create and activate a conda environment\
conda create -n nanopore_env python=3.9 -y
conda activate nanopore_env

 # Install NanoFilt, NanoStat, NanoPlot (python-based tools)
pip install NanoFilt NanoPlot NanoStat

# Install Porechop
pip install porechop 
#if porechop doesnt install, use the following as alternative
pip install cutadapt

#Download and compile Minimap2:
# Download Minimap2
git clone https://github.com/lh3/minimap2.git
cd minimap2
make
# Add minimap2 to PATH (optional)
export PATH=$PATH:$(pwd)

```

#STEP 1: Quality Control
1. Perform quality checks on fastq files (e.g., length distribution, quality scores).
2. Filter out low-quality reads using `NanoFilt` 
```{r}
#for it, enter the folder where your reads folder is located, The for loop iterates through all .fastq files in the reads_extracted folder.
#!/bin/bash

# Create results directory if it doesn't exist
mkdir -p results

# Loop through each FASTQ file in the reads_extracted folder
for file in reads_extracted/*.fastq; do
    # Extract the base name of the file (e.g., sample1 from sample1.fastq)
    base=$(basename "$file" .fastq)

    # Run NanoFilt for each file and save the output in the results directory
    NanoFilt -q 9 -l 1000 "$file" > "results/${base}_filtered.fastq"

    echo "Processed $file -> results/${base}_filtered.fastq"
done

#the filtered fastq reads will be saved in "results" folder.
```

#STEP 2: Activate QIIME2 and Import data


```{r pressure, echo=FALSE}


# Import feature table
qiime tools import \
  --input-path features/raw_feature_table.biom \
  --type 'FeatureTable[Frequency]' \
  --input-format BIOMV210Format \
  --output-path output/feature_table.qza

# Import metadata
cp metadata/metadata_table metadata.tsv
qiime metadata tabulate \
  --m-input-file metadata/metadata_table \
  --o-visualization output/metadata.qzv

#before importing the reads, rename and compress files as the further steps ahead require the specific pattern 

#Rename Files
# Navigate to the directory containing filtered files
cd results

#if using "???" Rename each file to match QIIME2's required format
for file in *_filtered.fastq; do
    sample=$(basename "$file" _filtered.fastq)
    mv "$file" "${sample}_L001_R1_001.fastq"
done

#compress the file
gzip results/*.fastq

#compress without name change
for file in results/*_filtered.fastq; do [[ -f "$file" ]] && gzip "$file" && echo "Compressed: $file"; done

##potential errors:
###1. If the files seems incorrectly zipped, you can check it by following: 
file results/* 
expected output: 
barcode01_L001_R1_001.fastq.gz: gzip compressed data
barcode02_L001_R1_001.fastq.gz: gzip compressed data
and,
ls results/
expected output:
barcode01_L001_R1_001.fastq.gz
barcode02_L001_R1_001.fastq.gz

###2. to see if QIIME is locating file directly:
find $PWD/results -name "*.fastq.gz"

###3. If any unrelated files exist, remove them:
rm results/unwanted_file

###4. create a manifest
nano results/manifest.csv

!/bin/bash

Set the output file name
manifest_file="results/manifest.csv"

Write the header line to the manifest file
echo "sample-id,absolute-filepath,direction" > "$manifest_file"

Loop through all .fastq.gz files in the results directory
for file in results/*.fastq.gz; do
    # Extract the sample ID from the filename
    sample_id=$(basename "$file" | cut -d_ -f1)
    
    # Append the sample details to the manifest file
    echo "$sample_id,$PWD/$file,forward" >> "$manifest_file"
done

echo "Manifest file created: $manifest_file"


Save the script as create_manifest.sh 
chmod +x create_manifest.sh
./create_manifest.sh

###5. if file is comma-separated (.csv), convert it into tab-separated (.tsv)
sed 's/,/\t/g' results/manifest.csv > results/manifest.tsv
verify the new manifest file
cat -T results/manifest.tsv

expected outcome:
sample-id^Iabsolute-filepath^Idirection
barcode01^I/home/ayesha/nanopore/results/barcode01_L001_R1_001.fastq.gz^Iforward

# Import FASTQ files as single-end reads (if these are barcoded single reads)
# For PHRED+33:
qiime tools import \
  --type 'SampleData[SequencesWithQuality]' \
  --input-path results/manifest.tsv \
  --output-path demux.qza \
  --input-format SingleEndFastqManifestPhred33V2

# For PHRED+64:
qiime tools import \
  --type 'SampleData[SequencesWithQuality]' \
  --input-path results/manifest.tsv \
  --output-path demux.qza \
  --input-format SingleEndFastqManifestPhred64V2

# Summarize sequence quality
qiime demux summarize \
  --i-data output/demux.qza \
  --o-visualization output/demux.qzv
```

#STEP 3: Denoising with DADA2
Assess and filter the sequencing reads.
```{r}

# Denoise and filter reads (e.g., using DADA2)
qiime dada2 denoise-single \
  --i-demultiplexed-seqs output/demux.qza \
  --p-trim-left 0 \
  --p-trunc-len 300 \
  --o-representative-sequences output/rep-seqs.qza \
  --o-table output/filtered-table.qza \
  --o-denoising-stats output/denoising-stats.qza

```

##Assign taxonomy
Classify features using a pre-trained classifier (e.g., SILVA, Greengenes).
```{r}
# Import taxonomy reference database
qiime tools import \
  --type 'FeatureData[Taxonomy]' \
  --input-path taxonomy/taxonomy_table \
  --output-path output/taxonomy.qza

# Assign taxonomy to features
qiime feature-classifier classify-sklearn \
  --i-classifier taxonomy/taxonomy.qza \
  --i-reads output/rep-seqs.qza \
  --o-classification output/taxonomy-assigned.qza

# Visualize taxonomy
qiime metadata tabulate \
  --m-input-file output/taxonomy-assigned.qza \
  --o-visualization output/taxonomy.qzv

```

##Generate Phylogenetic tree
```{r}
qiime phylogeny align-to-tree-mafft-fasttree \
  --i-sequences output/rep-seqs.qza \
  --o-alignment output/aligned-rep-seqs.qza \
  --o-masked-alignment output/masked-aligned-rep-seqs.qza \
  --o-tree output/unrooted-tree.qza \
  --o-rooted-tree output/rooted-tree.qza

```

##Diversity Analysis
```{r}
# Core metrics
qiime diversity core-metrics-phylogenetic \
  --i-phylogeny output/rooted-tree.qza \
  --i-table output/filtered-table.qza \
  --p-sampling-depth 1000 \
  --m-metadata-file metadata/metadata_table \
  --output-dir output/core-metrics-results

# Visualize alpha diversity
qiime diversity alpha-group-significance \
  --i-alpha-diversity output/core-metrics-results/faith_pd_vector.qza \
  --m-metadata-file metadata/metadata_table \
  --o-visualization output/faith-pd-significance.qzv

# Visualize beta diversity
qiime diversity beta-group-significance \
  --i-distance-matrix output/core-metrics-results/unweighted_unifrac_distance_matrix.qza \
  --m-metadata-file metadata/metadata_table \
  --o-visualization output/unweighted-unifrac-significance.qzv

```

##Cretae taxonomy barplots
```{r}
qiime taxa barplot \
  --i-table output/filtered-table.qza \
  --i-taxonomy output/taxonomy-assigned.qza \
  --m-metadata-file metadata/metadata_table \
  --o-visualization output/taxa-bar-plots.qzv

```

##Export data for downstream analysis
```{r}
qiime tools export \
  --input-path output/filtered-table.qza \
  --output-path export/

qiime tools export \
  --input-path output/taxonomy-assigned.qza \
  --output-path export/

```

#Install and Compile SAMtools, HTSlib, and BCFtools
These tools remain useful for working with aligned and processed data.

```{r}
sudo apt install autoconf
git clone https://github.com/samtools/htslib.git
git clone https://github.com/samtools/samtools.git
git clone https://github.com/samtools/bcftools.git

# HTSlib Installation
cd htslib
git submodule update --init --recursive
autoreconf -i
./configure
make
cd ..

# SAMtools Installation
cd samtools
autoheader
autoconf -Wno-syntax
./configure
make
cd ..

# BCFtools Installation
cd bcftools
autoheader
autoconf -Wno-syntax
./configure
make

#Minimap installation
git clone https://github.com/lh3/minimap2.git
cd minimap2 cd minimap2 && make && sudo cp minimap2 /usr/local/bin/
cd ..

#Install Picard for File Manipulations
git clone https://github.com/broadinstitute/picard.git
cd picard
git checkout 5db8017
./gradlew shadowJar

#Install GATK for Variant Calling
#GATK can still be used for variant calling with Nanopore reads.
wget https://github.com/broadinstitute/gatk/releases/download/4.2.6.1/gatk-4.2.6.1.zip
unzip gatk-4.2.6.1.zip

#Install SRA Toolkit
wget https://ftp-trace.ncbi.nlm.nih.gov/sra/sdk/3.1.1/sratoolkit.3.1.1-ubuntu64.tar.gz
mkdir sratoolkit
tar -xzf sratoolkit.3.1.1-ubuntu64.tar.gz -C sratoolkit
./sratoolkit/sratoolkit.3.1.1-ubuntu64/bin/vdb-config --interactive
./sratoolkit/sratoolkit.3.1.1-ubuntu64/bin/fasterq-dump

# Install Porechop (for adapter trimming)
git clone https://github.com/rrwick/Porechop.git
cd Porechop && sudo python3 setup.py install

# Install Medaka (variant calling for nanopore data)
pip install medaka


# Install Nanopolish (for methylation/base modification analysis)
git clone https://github.com/jts/nanopolish.git
cd nanopolish && make && sudo cp nanopolish /usr/local/bin/

# Install pycoQC (for nanopore sequencing quality control)
pip install pycoQC

```

#Create a bin Directory and Set Up Symlinks

```{r}
mkdir nbin
cd nbin
ln -s /home/ayesha/software/minimap2/minimap2 minimap2
ln -s /home/ayesha/software/htslib/bgzip bgzip
ln -s /home/ayesha/software/samtools/samtools samtools
ln -s /home/ayesha/software/bcftools/bcftools bcftools
ln -s /home/ayesha/software/sratoolkit/sratoolkit.3.1.1-ubuntu64/bin/fasterq-dump fasterq-dump
ln -s /home/ayesha/software/nanofilt/nanofilt nanofilt
ls  # Verify the symlinks

#Add bin Directory to PATH in .bashrc
nano ~/.bashrc

# Add at the end:
export PATH="$HOME/abin:$PATH"

# Reload .bashrc
source ~/.bashrc

```

#Set Up the Workspace

```{r}
mkdir nanopore_pipeline
cd nanopore_pipeline

# Create directories for organization
mkdir reads taxonomy metadata features output
# Move your data into the appropriate folders
mv /path_to_your_data/*.fastq reads/
mv /path_to_your_data/taxonomy_table.tsv taxonomy/
mv /path_to_your_data/metadata.csv metadata/
mv /path_to_your_data/raw_feature_table.tsv features/

```

#Quality Filtering
Filter the reads for quality using NanoFilt:

```{r}
cd reads
nanofilt -q 7 < input_reads.fastq > filtered_reads.fastq

```

#Taxonomic Classification

```{r}
#Import your data into QIIME2:
qiime tools import \
    --type 'SampleData[SequencesWithQuality]' \
    --input-path filtered_reads.fastq \
    --output-path demux.qza

#Perform taxonomic classification:
qiime feature-classifier classify-sklearn \
    --i-classifier /path_to_pretrained_classifier.qza \
    --i-reads demux.qza \
    --o-classification taxonomy.qza

#Export the results:
qiime tools export --input-path taxonomy.qza --output-path taxonomy/

```

#Alignment (Minimap2)

```{r}
minimap2 -ax map-ont reference_genome.fa filtered_reads.fastq > aligned.sam
samtools view -bS aligned.sam > aligned.bam
samtools sort aligned.bam -o aligned.sorted.bam
samtools index aligned.sorted.bam

```

#Generate Feature Tables
If your raw feature table and taxonomy data need updating:
Use QIIME2’s feature table methods

```{r}
qiime feature-table summarize \
    --i-table feature_table.qza \
    --o-visualization feature_table_summary.qzv
```

#Metadata Analysis
Incorporate your metadata for diversity or statistical analysis:

```{r}
qiime diversity core-metrics-phylogenetic \
    --i-table feature_table.qza \
    --m-metadata-file metadata/metadata.csv \
    --output-dir core_metrics_results

```

#Visualization in QIIME2

```{r}
#1. Taxonomy Bar Plots
qiime taxa barplot \
    --i-table feature_table.qza \
    --i-taxonomy taxonomy.qza \
    --m-metadata-file metadata/metadata.csv \
    --o-visualization taxonomy_barplot.qzv

qiime tools view taxonomy_barplot.qzv

#2. Alpha and Beta Diversity
#Generate Diversity Metrics:
qiime diversity core-metrics-phylogenetic \
    --i-table feature_table.qza \
    --i-phylogeny rooted-tree.qza \
    --m-metadata-file metadata/metadata.csv \
    --p-sampling-depth 10000 \
    --output-dir core_metrics_results
#Visualize Alpha Diversity:
qiime diversity alpha-group-significance \
    --i-alpha-diversity core_metrics_results/faith_pd_vector.qza \
    --m-metadata-file metadata/metadata.csv \
    --o-visualization alpha_diversity_significance.qzv
#Visualize Beta Diversity:
qiime diversity beta-group-significance \
    --i-distance-matrix core_metrics_results/unweighted_unifrac_distance_matrix.qza \
    --m-metadata-file metadata/metadata.csv \
    --m-metadata-column sample_type \
    --o-visualization beta_diversity_significance.qzv \
    --p-pairwise
#View the .qzv files:
qiime tools view alpha_diversity_significance.qzv
qiime tools view beta_diversity_significance.qzv

```

